Estimating the Financial Impact of Wildfire Smoke on US National parks

### 2020-05-20 Markdown Documents/Manuscript.Rmd
Rough draft of intro done

This is how you add a reference (e.g., [@fisichelli_protected_2015; @fann_ch_2016])

Points:
1. Public lands are important
2. NPs are specifically high value
3. NPs are under funded. This under funding prevents them from doing those important jobs
4. NPs get a lot of their funding from visitation. Vis is influenced by a lot of things. One of these things that's particularly out of their control is smoke from adjacent forested areas.
5. Smoke is bad for recreation and should decrease vis.
6. Smoke events have been increasing over the last 50 years. Particularly, these increases have been in the west.
7. These smoke events are expected to worsen via climate change over the next 50 years.
8. Logically, smoke should influence visitation to parks. This decrease should affect income. We seek to assess this effect in a systematic way and backcast what the monetary effects might have been.
9. Given the trends we observe, we project out into the future what the effects might be over the next 50 years.

Twenty eight percent of the land area of the United States is designated as public land [@vincent_federal_2020]. These public lands are the critical backdrop for the majority wildlife conservation and outdoor recreation in the country [@romagosa_inside_2015; @leemans_millennium_2003; @dudley_national_2011]. In a 2016 study, 93% of respondents indicated that they believe public open spaces and recreational areas should be protected, regardless of whether or not they visited them personally [@haefele_total_2016]. The crown jewel of the United States Public lands system is the National Park Service. The national park system alone received approximately 318 million recreational visitors and generated over $40 billion in economic output to local economies in 2018 [@thomas_2018_2018].

National parks undeniably represent a critical infrastructure for supporting ecological integrity and outdoor recreation in the United States, but have recently drawn public attention for considerable under funding [@simmonds_crisis_2018;@duncan_opinion_2016]. Most notably for the $12 billion repair backlog for basic maintenance and resource protection [@nps_nps_2018]. In addition to necessary repairs, the National Park Service reports a 14% reduction in staff funding over the last 5 years [@doi_fiscal_2019]. While the majority of National Park Service funding comes from congressional appropriations, recent budgetary restrictions have put pressure on parks to gather an increasing amount of their budget from recreational visitor fees such as admissions and camping.  The income from these fees has been steadily increasing over the last 10 years as parks have seen visitation increases while simultaneously increasing visitor fees [@gao_national_2016].

As national parks become increasingly dependent on visitor fees in order to provide the suite of social and ecological services communities depend on, it is of paralleled importance to understand the drivers of fluctuation in recreational visits. Research has shown that economic considerations (e.g. park fees) are considerably less influential on park visitation than are recreational opportunities [@neuvonen_visits_2010;@stevens_declining_2014]. Most factors which influence the recreational accessibility of a given park are either fixed (natural resources within the park area) or within the control of park management (permitting, trail building, etc.). Wildfire smoke however, shows considerable yearly variation and is largely produced by adjacent public lands and therefore out of the jurisdiction of park management [@dombeck_wildfire_2004;@collins_regional_2006].

>>>>

Twenty eight percent of the land area of the United States is designated as public land [@vincent_federal_2020]. These public lands are the critical backdrop for the majority wildlife conservation and outdoor recreation in the country [@romagosa_inside_2015; @leemans_millennium_2003; @dudley_national_2011]. In a 2016 study, 93% of respondents indicated that they believe public open spaces and recreational areas should be protected, regardless of whether or not they visited them personally [@haefele_total_2016]. The crown jewel of the United States public lands system is the National Park Service. The national park system alone received approximately 318 million recreational visitors and generated over $40 billion in economic output to local economies in 2018 [@thomas_2018_2018].

National parks undeniably represent a critical infrastructure for supporting ecological integrity and outdoor recreation in the United States, but have recently drawn public attention for considerable under funding [@simmonds_crisis_2018;@duncan_opinion_2016]. Most notably for the $12 billion repair backlog for basic maintenance and resource protection [@nps_nps_2018]. In addition to necessary repairs, the National Park Service reports a 14% reduction in staff funding over the last 5 years [@doi_fiscal_2019]. While the majority of National Park Service funding comes from congressional appropriations, recent budgetary restrictions have put pressure on parks to gather an increasing amount of their budget from recreational visitor fees such as admissions and camping. The income from these fees has steadily increased over the last 10 years as park visitation and visitor fees have both surged [@gao_national_2016].

As national parks become increasingly dependent on visitor fees in order to provide the suite of social and ecological services communities depend on, it is of paralleled importance to understand what drives fluctuation in recreational visits. Research has shown that economic considerations (e.g. park fees) are considerably less influential on park visitation than are recreational opportunities [@neuvonen_visits_2010;@stevens_declining_2014]. Most factors which influence the recreational accessibility of a given park are either fixed (natural resources within the park area) or within the control of park management (permitting, trail building, etc.). Wildfire smoke however, shows considerable yearly variation and is largely produced by adjacent public lands and therefore out of the jurisdiction of park management [@dombeck_wildfire_2004;@collins_regional_2006].

Wildfire smoke is a mixture of many gasses aerated by biomass combustion across a landscape [@naeher_woodsmoke_2007]. While many of these compounds have been shown to cause health problems in humans, organic carbon and black carbon particulate matter less than 2.5 μm in diameter (PM2.5) is particularly harmful to human health, causing between 260,000 - 600,000 global deaths annually [@johnston_fay_h_estimated_2012; @naeher_woodsmoke_2007]. Health concerns regarding wildfire smoke, specifically PM2.5, regurlarly drive calls for individuals to reduce outdoor recreation behavior [@oneill_summary_2013].

Wildfires and the smoke they produce have been increasing in the United States over the last 50 years, specifically in the American West and Inter-Mountain regions [@kitzberger_contingent_2007;@westerling_increasing_2016]. These regions are also projected to experience increased wildfire frequency and severity over the next 50 years as a result of climate change [@harvey_human-caused_2016; @peterson_new_2003; @barbero_climate_2015]. Anecdotally, it is known that wildfire smoke has impacted the recreation behavior of individuals in these areas and decreased visitation revenue to the national parks in both regions. This assumption however has not yet been demonstrated empirically.

The primary goal of this project is to assess how visitation to national parks in the Inter-Mountain and Pacific regions has systematically varied with surface level concentrations of PM2.5 resulting from wildfire smoke from 1980 - 2018. Given the variability in recreational fees between parks and over time, the secondary goal of this project is to produce order of magnitude estimates of the financial losses that wildfire smoke has caused  park units during the same time period. The intention of this research is not to perform an economic analysis of smoke effects on each park specifically, but rather to provide an approximation of how wildfire smoke affects population level recreational decision making and what the downstream economic impacts have been historically and may be in the future.
